Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2022-09-05 paper/paper-infer.tex
save

@@

Most traits and diseases are heritable, if not all. 
What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
Some phenotypes, such as height and schizophrenia are highly heritable [TODO: ADD CITATIONS]. 
These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them [TODO: ADD CITATIONS].
Knowing how much a phenotype is heritable and polygenic can teach a lot about it.
Therefore, many methods have been developed to estimate the (SNP) heritability $h^2$ and polygenicity $p$.
These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,zeng2021widespread,johnson2021estimation}.

These two parameters are useful e.g.\ to determine how well we can predict a phenotype from using genetic variants alone [TODO: ADD CITATIONS]. 
Such genetic predictors are called polygenic scores (PGS), and are getting closer to be included as part of existing clinical risk models for diseases [TODO: ADD CITATIONS].
LDpred2 is a very competitive polygenic score method that can directly builds PGS using summary statistics results from genome-wide associations studies (GWAS), making it highly applicable \cite[]{prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
LDpred2 uses the (SNP) heritability and polygenicity as parameters of its model. In LDpred2-auto, it can directly estimate these parameters from the data, making it applicable even when no validation data is available for tuning these two hyper-parameters of the model \cite[]{prive2020ldpred2}.

Here we extend LDpred2-auto to make it a highly reliable method for estimating (local) $h^2$, (per-variant) $p$, and $\alpha$.
We show that we can also reliably estimate the predictive ability of PGS derived from LDpred2-auto.

[TO FINISH]

[ALSO TALK ABOUT NEW SET OF VARIANTS]

@@

@@

For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}; we use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
We first simulate continuous phenotypes using function \texttt{snp\_simuPheno} from R package bigsnpr \cite[]{prive2017efficient}, varying three parameters: the SNP-heritability $h^2$, the polygenicity $p$ (i.e. the proportion of causal variants), and the parameter $\alpha$ in equation \eqref{eq:new_model} that controls the relationship between minor allele frequencies and expected effect sizes.
This function first picks a proportion $p$ of causal variants at random, sample effect sizes $\gamma$ using the variance component parameterized by $\alpha$ and then scale the effect sizes so that the genetic component $G \gamma$ has a variance $h^2$, where $G$ is the genotype matrix. Finally, some Gaussian noise is added so that the full phenotype has a variance of 1. 
Then, a GWAS is run to obtain summary statistics using $N$ (either the 200,000 dedicated to this, or a subset of 20,000) individuals using a simple linear regression implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
Finally, we run the new LDpred2-auto model (Methods) with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled). 
LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$.

LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha}).
Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figure \ref{fig:simu_h2}).
When power is low, the polygenicity can be overestimated when the true value is very small (e.g.\ $p=0.0005$), and underestimated when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
The $\alpha$ estimate becomes unreliable when power is too low, which can be detected by a small number of chains kept from LDpred2-auto (Figure \ref{fig:simu_alpha}).
Moreover, LDpred2-auto can also infer per-variant probabilities of being causal and per-block (local) heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
Finally, LDpred2-auto can be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler, even when power is low (Figure \ref{fig:simu_r2}).

We also run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a thresholding corresponding to the prevalence.
Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$). The only difference is the transformation to the liability scale that needs to be applied to the $h^2$ and $r^2$ estimates from LDpred2-auto, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ \cite[]{grotzinger2021pervasive}.

@@

Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a smaller polygenicity (Figure \ref{fig:ukbb_h2}).
Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure \ref{fig:ukbb_h2_p}).
Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}), which is consistent with widespread negative selection.
As for the inferred predictive performance, they are highly consistent with the ones derived from the test; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}). Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$, and these are higher than values reported in the literature.

We then investigate previous estimates using two alternative LD references (Methods).
Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:alt_LD}).

We then investigate previous estimates using the extended set of HapMap3 variants proposed here (Methods). As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [10.8-13.7]) and lower $p$ (decrease of 11.5\% [10.7-12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}). 
This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1-8.2]).
A particularly much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502-0.514] instead of 0.324 [0.320-0.329]), which is also reflected in a larger predictive performance ($r^2$ in test set of 0.516 [0.508-0.524] instead of 0.344 [0.335-0.353]).
Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}).

Finally, we investigate different transformations to apply to some continuous phenotypes used here.
Indeed, 49 of such phenotypes seems log-normally distributed or heavy-tailed (when visualizing their histogram); we therefore log-transform them.
However, we do investigate alternative transformations here to decide which one should be preferred and to check the impact on the inference from LDpred2-auto.
Note that we use the HapMap3+ set of variants here.
We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates for $p$ and $\alpha$ are highly consistent.
However, $h^2$ estimates and predictive performance $r^2$ (in the test set) are generally larger with the log-transformation, hinting that it probably makes sense to transform these phenotypes.
We then compare to using the rank-based inverse normal (RIN) transformation in Figure \ref{fig:RINT}; estimates for $p$ and $\alpha$ are also highly consistent.
Except for bilirubin and lipoprotein(a) concentration, generally higher $h^2$ estimates and predictive performance $r^2$ are obtained with the RIN-transformation than the log-transformation.

@@

Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks in Figure \ref{fig:median_local_h2}. The median heritability explained by a block of variants is largely proportional to the number of variants in this block.
Unsurprisingly, the outlier block explaining a much larger heritability contains the HLA region (Figure \ref{fig:median_local_h2}).
Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
Some variants in multiple small regions across the genome seem to have a much larger probability of being causal; these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.

@@

>>>>

	TODO

@@

Most traits and diseases are heritable, if not all. 
What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
Some phenotypes, such as schizophrenia or height, are highly heritable \cite[]{sullivan2003schizophrenia,yang2010common}. 
These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
Knowing how much a phenotype is heritable and polygenic can teach a lot about it.
Therefore, many methods have been developed to estimate the (SNP) heritability $h^2$ and polygenicity $p$.
These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,zeng2021widespread,johnson2021estimation}.

These two parameters are useful e.g.\ to determine how well we can predict a phenotype from using genetic variants alone \cite[]{daetwyler2008accuracy,dudbridge2013power}. 
Such genetic predictors are called polygenic scores (PGS), and are getting closer to be included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
LDpred2 is a very competitive polygenic score method that can directly builds PGS using summary statistics results from genome-wide associations studies (GWAS), making it highly applicable \cite[]{prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
LDpred2 uses the (SNP) heritability and polygenicity as parameters of its model. In LDpred2-auto, it can directly estimate these parameters from the data, making it applicable even when no validation data is available for tuning these two model hyper-parameters \cite[]{prive2020ldpred2}.

Here we extend LDpred2-auto to make it a highly reliable method for estimating (local) $h^2$, (per-variant) $p$, and $\alpha$.
So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS (without needing an independent test set).
Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$.
We call this new preferred set of 1,444,196 SNPs HapMap3+.

@@

[TODO: CHOOSE FIGURES TO HAVE IN MAIN TEXT]

@@

For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}; we use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
We first simulate continuous phenotypes using function \texttt{snp\_simuPheno} from R package bigsnpr \cite[]{prive2017efficient}, varying three parameters: the SNP-heritability $h^2$, the polygenicity $p$ (i.e. the proportion of causal variants), and the parameter $\alpha$ in equation \eqref{eq:new_model} that controls the relationship between minor allele frequencies and expected effect sizes.
This function first picks a proportion $p$ of causal variants at random, sample effect sizes $\gamma$ using the variance component parameterized by $\alpha$ and then scale the effect sizes so that the genetic component $G \gamma$ has a variance $h^2$, where $G$ is the genotype matrix. Finally, some Gaussian noise is added so that the full phenotype has a variance of 1. 
Then, a GWAS is run to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a subset of 20,000) using a simple linear regression implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
Finally, we run the new LDpred2-auto model (Methods) with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled). 
LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$.

LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha}).
Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figure \ref{fig:simu_h2}).
When power is low, the polygenicity can be overestimated when the true value is very small (e.g.\ $p=0.0005$), and underestimated when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
The $\alpha$ estimate becomes unreliable when power is too low, which can be detected by a small number of chains kept from LDpred2-auto (Figure \ref{fig:simu_alpha}).
Moreover, LDpred2-auto can also infer per-variant probabilities of being causal and per-block (local) heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler, even when power is low (Figure \ref{fig:simu_r2}).

We then run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a thresholding corresponding to the prevalence.
Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$). The only difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2012better}, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ in LDpred2-auto or LD Score regression \cite[]{grotzinger2021pervasive}.

@@

Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a smaller polygenicity (Figure \ref{fig:ukbb_h2}).
Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure \ref{fig:ukbb_h2_p}).
Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}), which is consistent with widespread negative selection.
As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}). Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$, and because these estimates are higher than values reported in the literature.

We then investigate using two alternative LD references (Methods).
Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:alt_LD}).

We then investigate using the extended set of HapMap3 variants proposed here (Methods). As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [10.8-13.7]) and lower $p$ (decrease of 11.5\% [10.7-12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}). 
This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1-8.2]).
A particularly much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502-0.514] instead of 0.324 [0.320-0.329]), which is also reflected in a larger predictive performance ($r^2$ in test set of 0.516 [0.508-0.524] instead of 0.344 [0.335-0.353]).
Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}).

Finally, we investigate different transformations to apply to some continuous phenotypes used here.
Indeed, 49 of such phenotypes seems log-normally distributed or heavy-tailed (when visualizing their histogram); we therefore log-transform them.
However, we do investigate alternative transformations here to decide which one should be preferred and to check how this impacts the inference from LDpred2-auto.
Note that we now use the HapMap3+ set of variants here.
We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates of $p$ and $\alpha$ are highly consistent.
However, $h^2$ estimates and predictive performance $r^2$ (in the test set) are generally larger with the log-transformation, hinting that it probably makes sense to transform these phenotypes.
We then compare to using the rank-based inverse normal (RIN) transformation in Figure \ref{fig:RINT}; estimates for $p$ and $\alpha$ are also highly consistent.
Except for bilirubin and lipoprotein(a) concentration, generally higher $h^2$ estimates and predictive performance $r^2$ are obtained with the RIN-transformation than the log-transformation.

@@

Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks; the median heritability explained by a block of variants is largely proportional to the number of variants in this block (Figure \ref{fig:median_local_h2}).
The outlier block explaining a much larger heritability contains the HLA region.
Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.

@@

[MENTION PARTIAL COR?]
